---
title: "ADT QC Report"
author: "Zach Thomson"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    code_folding: hide
    df_print: paged
    self_contained: true
params: 
  in_counts: NULL
  in_bcs: NULL
  well_id: NULL
  out_dir: NULL
---

<a id="contents"></a>

## Contents

#### [Data Processing](#data_processing)
- [Session Preparation](#session_preparation)
- [Load Inputs](#load_inputs)
- [Create Seurat Object](#seurat_object)
- [Create ADT Dataframe](#adt_data_frame)

#### [QC Plots](#plots)
- [ADT Cells vs Background](#split_violin)
- [ADT UMIs/Cell Stats](#adt_stats)
- [Total Counts by ADT](#total_counts)
- [ADT UMAP](#adt_umap)

#### [Session Info](#session_info)

<a id="data_processing"></a>

## Data Processing

<a id="session_preparation"></a>

### Session Preparation

#### Load Libraries
```{r load libraries}
start_time <- Sys.time()

quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}
quiet_library(ggplot2)
quiet_library(Seurat)
quiet_library(viridis)
quiet_library(dplyr)
quiet_library(plyr)
quiet_library(jsonlite)
quiet_library(H5weaver)
```

Declaring start
```{r Declare Start}
stm("Starting ADT Report QC")
```

[Return to Contents](#contents)

####Check Input Files
```{r check inputs}
if(!file.exists(params$in_counts)){
  stm(paste0("ERROR: Cannot find IN COUNTS:", in_counts))
  stop()
}
if(!file.exists(params$in_bcs)){
  stm(paste0("ERROR: Cannot find IN BARCODES", in_bcs))
  stop()
}
```

<a id="load_inputs"></a>

### Load Inputs

#### Load Input Files
```{r locate input files}
stm(paste0("Reading input files"))

adt_count_mtx <- read.csv(params$in_counts)
cell_barcodes <- read.table(params$in_bcs, sep = '\t')
cell_barcodes <- gsub('.{2}$', '', cell_barcodes$V1)
well_id <- params$well_id
out_dir <- params$out_dir
```

#### Create out directory if missing
```{r Create Out Dir}
if(!dir.exists(out_dir)) {
  stm(paste0("Creating Output Directory: ",out_dir))
  dir.create(out_dir, 
             recursive = TRUE)
}
```

<a id="seurat_object"></a>

#### Create Seurat Object
```{r Create Seurat Object}
stm(paste0("Creating Seurat Object"))

diff_bcs <- setdiff(adt_count_mtx$cell_barcode, cell_barcodes)
neg <- subset(adt_count_mtx, subset = adt_count_mtx$cell_barcode %in% diff_bcs)
neg_50k <- neg[order(neg$total, decreasing = T),]
neg_50k <- neg_50k[1:50000,]
rownames(neg_50k) <- neg_50k$cell_barcode

pos_subset <- subset(adt_count_mtx, subset = adt_count_mtx$cell_barcode %in% cell_barcodes)
rownames(pos_subset) <- pos_subset$cell_barcode

pos_so <- CreateSeuratObject(counts = t(pos_subset[3:length(colnames(pos_subset))]), project = 'Cells')
neg_so <- CreateSeuratObject(counts = t(neg_50k[3:length(colnames(neg_50k))]), project = 'Top 50k Empty Droplets')
so_merge <- merge(pos_so, neg_so)
so_merge[['well']] <- rep(well_id, length(so_merge@assays$RNA))

```

#### UMAP clustering for ADTs
```{r UMAP Clustering for ADTs}
stm(paste0("Clustering ADTs"))
so_merge <- SetIdent(so_merge, value = 'orig.ident')
adt_positives <- subset(so_merge, idents = 'Cells')
VariableFeatures(adt_positives) <- rownames(adt_positives)

if (length(rownames(adt_positives)) > 30) {
  dims = 30
} else {
  dims = length(rownames(adt_positives)) - 1
}

adt_positives <- NormalizeData(adt_positives, normalization.method = 'CLR', margin = 2) %>% 
  ScaleData() %>% RunPCA(npcs = dims)
adt_positives <- RunUMAP(adt_positives, dims = 1:(dims)) %>% FindNeighbors() %>% FindClusters()
```

<a id="adt_data_frame"></a>

#### ADT UMIs Dataframe construction
```{r ADT UMIs dataframe construction}
stm(paste0("Summing ADTs"))
adt_subset <- adt_count_mtx[,3:length(colnames(adt_count_mtx))]
adt_sums <- data.frame(colSums(adt_subset))
adt_sums$adt <- rownames(adt_sums)
colnames(adt_sums) <- c('Count', 'ADT')

pos_adts <- pos_subset[,3:length(colnames(pos_subset))]
pos_adt_sums <- data.frame(colSums(pos_adts))
pos_adt_sums$adt <- rownames(pos_adt_sums)
pos_adt_sums$Cell <- rep('True', length(pos_adt_sums$adt))
colnames(pos_adt_sums) <- c('Count','ADT','Cell')

neg_equal <- neg[order(neg$total, decreasing = T),]
neg_equal <- neg_equal[1:length(rownames(pos_subset)),]
rownames(neg_equal) <- neg_equal$cell_barcode

neg_adts <- neg_equal[,3:length(colnames(neg_equal))]
neg_adt_sums <- data.frame(colSums(neg_adts))
neg_adt_sums$adt <- rownames(neg_adt_sums)
neg_adt_sums$Cell <- rep('Background', length(neg_adt_sums$adt))
colnames(neg_adt_sums) <- c('Count','ADT','Cell')

combined_adts <- rbind(pos_adt_sums, neg_adt_sums)
                           
```

[Return to Contents](#contents)

<a id="plots"></a>
### QC Plots

<a id="split_violin"></a>
#### Split Violin Plot
```{r split violin plot, fig.width=10, fig.height=8}
VlnPlot(so_merge, features = 'nCount_RNA', log = T, 
        pt.size = 0, split.plot = T, split.by = 'orig.ident', group.by = 'well') & geom_hline(yintercept=1000, linetype="dashed", color = "blue") &
  geom_hline(yintercept=300, linetype="solid", color = "red") & stat_summary(fun=median, geom="point", size=2, color="green") & ggtitle('ADT UMIs/Cell') &
  theme(axis.title.x=element_blank())
```

<a id="adt_stats"></a>

#### ADT UMIs/Cell Stats
```{r ADT UMIs/Cell Stats}
overall_stats<- data.frame("Total")
overall_stats$pos_adt <- median(pos_subset$total)
overall_stats$neg_adt <- median(neg_50k$total)
overall_stats$signal_ratio <- (overall_stats$pos_adt / overall_stats$neg_adt)
colnames(overall_stats) <-  c('Marker','Median Cells','Median Background','Signal Ratio')
overall_stats$`Signal Ratio` <- round(overall_stats$`Signal Ratio`, 2)

adt_stats <- data.frame(colnames(pos_adts))
adt_stats$pos_adts <- apply(pos_adts, 2, median)
adt_stats$neg_adts <- apply(neg_adts, 2, median)
adt_stats$signal_ratio <- (adt_stats$pos_adts / adt_stats$neg_adts)
colnames(adt_stats) <- c('Marker','Median Cells','Median Background','Signal Ratio')
adt_stats$`Signal Ratio` <- round(adt_stats$`Signal Ratio`, 2)

final_table <- rbind(overall_stats, adt_stats)

qc_table(final_table)
```

<a id="total_counts"></a>

#### ADT UMIs plot - Linear Scale
```{r ADT UMIs plot, fig.width=10, fig.height=6}
ggplot(combined_adts, aes(x = reorder(ADT, -Count), y = Count, color = Cell)) + geom_point(size = 5) + 
  geom_hline(yintercept=median(adt_sums$Count), linetype="dashed", color = "blue") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))# + scale_y_log10()
```

#### ADT UMIs plot - Log Scale
```{r ADT UMIs plot log, fig.width=10, fig.height=6}
ggplot(combined_adts, aes(x = reorder(ADT, -Count), y = Count, color = Cell)) + geom_point(size = 5) + 
  geom_hline(yintercept=median(adt_sums$Count), linetype="dashed", color = "blue") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + scale_y_log10()
```


<a id="adt_umap"></a>

```{r dynamically scale adt umap height}
n_features <- length(rownames(adt_positives))
height = ceiling((n_features / 3) * 2)
```

#### UMAP clustering for ADTs individually
```{r UMAP clustering for ADTs individually, fig.width=10, fig.height=height, message=F, echo=F}
FeaturePlot(adt_positives, features = rownames(adt_positives), min.cutoff = 'q03', max.cutoff = 'q97', label.size = 1, ncol = 3) & scale_color_viridis(option = 'inferno')
```

[Return to Contents](#contents)

#### Export Positive Droplet ADT Matrix & Metadata
```{r export adt matrix}
write.csv(pos_adts, file = file.path(out_dir,paste0(well_id,
                                                    "_adt_positive_tag_counts.csv")))
meta_data <- pos_subset[,1:2]
meta_data$flag <- ifelse(meta_data$total > 300, "Good", "Low")
rownames(meta_data) <- meta_data$cell_barcode

write.csv(meta_data, file = file.path(out_dir, paste0(well_id,
                                                      "_adt_metadata.csv")))
```

<a id="session_info"></a>

## Session Information

```{r Session Info}
sessionInfo()
```

Total time elapsed
```{r Show Time}
end_time <- Sys.time()
diff_time <- end_time - start_time
time_message <- paste0("Elapsed Time: ", 
                       round(diff_time, 3),
                       " ", units(diff_time))
print(time_message)
stm(time_message)
stm("ADT QC Report complete.")
```

[Return to Contents](#contents)
