---
title: "Split 10x Genomics scRNA-seq .h5 file based on HTOparser results"
author: 
 - Lucas Graybuck
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    df_print: paged
    self_contained: true
params:
  in_h5: NULL
  in_mol: NULL
  in_mat: NULL
  in_tbl: NULL
  in_well: NULL
  out_dir: NULL
---

```{r Load Libraries}
start_time <- Sys.time()

quiet_library <- function(...) {
  suppressPackageStartupMessages(library(...))
}
quiet_library(rhdf5)
quiet_library(H5weaver)
quiet_library(Matrix)
quiet_library(ggplot2)
quiet_library(cowplot)
```

Declaring start
```{r Declare start}
stm("Starting .h5 split")
```

Argument parsing
```{r Parse arguments}
if(is.null(params$in_h5)) {
  in_h5 <- system.file("testdata/well1.h5", package = "H5weaver")
  in_mol <- system.file("testdata/well1_molecule_info.h5", package = "H5weaver")
  in_mat <- system.file("testdata/well1_count_matrix.csv.gz", package = "H5weaver")
  in_tbl <- system.file("testdata/well1_category_table.csv.gz", package = "H5weaver")
  in_well <- "B000-P0C0W0"
  out_dir <- tempdir()
} else {
  in_h5 <- params$in_h5
  in_mol <- params$in_mol
  in_mat <- params$in_mat
  in_tbl <- params$in_tbl
  in_well <- params$in_well
  out_dir <- params$out_dir
}

stm(paste0("IN  H5 file         : ", in_h5))
stm(paste0("IN  Mol Info H5     : ", in_mol))
stm(paste0("IN  Count matrix    : ", in_mat))
stm(paste0("IN  Category table  : ", in_tbl))
stm(paste0("IN  Well ID         : ", in_well))
stm(paste0("OUT Directory       : ", out_dir))
```

```{r Print Arguments}
print(c(
  paste0("IN  H5 file         : ", in_h5),
  paste0("IN  Mol Info H5     : ", in_mol),
  paste0("IN  Count matrix    : ", in_mat),
  paste0("IN  Category table  : ", in_tbl),
  paste0("IN  Well ID         : ", in_well),
  paste0("OUT Directory       : ", out_dir)
))
```

Load HTO counts
```{r Load Count Matrix}
stm(paste0("Loading HTO count matrix from ", in_mat))
hash_count_df <- fread(in_mat)

hash_count_mat <- as.matrix(hash_count_df[,-1])

rownames(hash_count_mat) <- hash_count_df[[1]]
rm(hash_count_df)
```

Load HTO categories
```{r Load Category Table}
stm(paste0("Loading HTO category table from ", in_tbl))
hash_category_table <- fread(in_tbl)

hto_barcodes <- unique(hash_category_table$hto_barcode[hash_category_table$hto_category == "singlet"])
```

Display expected output files
```{r Display Outputs}
output_h5_files <- file.path(out_dir, paste0(in_well, "_", 
                                             c(hto_barcodes,"multiplet"),
                                             ".h5"))

output_messages <- paste0("OUT HDF5            : ", output_h5_files)

for(i in seq_along(output_messages)) {
  stm(output_messages[i])
}

print(output_messages)
```

Load scRNA_seq  Dataset
```{r Load Query}
stm(paste0("Loading HDF5 from ", in_h5))
h5_list <- h5dump(in_h5)
```

Read molecule info to get read counts per cell
```{r Assemble Read Counts}
stm(paste0("Assembling Read Counts per Cell from ", in_mol))

bc <- sub("-1","",h5_list$matrix$barcodes)

bc_counts <- data.table(mol_idx = h5read(in_mol, "/barcode_idx"),
                        umi_count = h5read(in_mol, "/count"))

bc_sums <- bc_counts[, .(n_reads = sum(umi_count)), by = mol_idx]
rm(bc_counts)

mol_bc <- h5read(in_mol, "/barcodes")
bc_sums$cell_barcode <- mol_bc[bc_sums$mol_idx + 1]
rm(mol_bc)

bc_sums <- bc_sums[,.(cell_barcode, n_reads)]

n_reads <- bc_sums$n_reads[match(bc, bc_sums$cell_barcode)]
n_reads[is.na(n_reads)] <- 0

if("observations" %in% names(h5_list$matrix)) {
  h5_list$matrix$observations$n_reads <- n_reads
} else {
  h5_list$matrix$observations <- list(n_reads = n_reads)
}
```

Compute N UMIs and N Genes per cell
```{r n_umi and n_genes}
stm("Computing UMI and Gene Counts per Cell")

h5_list <- h5_list_convert_to_dgCMatrix(h5_list, target = "matrix")
h5_list$matrix$observations$n_umis <- unname(colSums(h5_list$matrix_dgCMatrix))
h5_list$matrix$observations$n_genes <- unname(colSums(h5_list$matrix_dgCMatrix > 0))

h5_list <- h5_list_convert_from_dgCMatrix(h5_list, target = "matrix")
```

Add cell ids
```{r}
stm("Adding Cell UUIDs and Names")

h5_list <-add_cell_ids(h5_list,
                       add_uuid = TRUE,
                       replace_barcode = TRUE,
                       retain_original_barcode = TRUE,
                       add_name = TRUE)
```

Generate Chip, Pool, and Batch IDs based on well_id
```{r Chip Pool Batch}
stm("Adding Batch, Pool, Chip, and Well metadata")

h5_list <- add_well_metadata(h5_list,
                             well_id = in_well)
```

Add hash data to the list object
```{r Add Hash Data}
stm("Adding hash data to cellranger results")

h5_list <- add_h5_list_hash_results(h5_list,
                                    hash_category_table,
                                    hash_count_mat,
                                    match_target = "original_barcodes")
```

Extract metadata for plotting
```{r Assemble Metadata}
stm("Generating tables and plots for report")
meta <- h5_list_cell_metadata(h5_list)
```

Show read/umi/gene stats overall
```{r Read UMI Gene Stats}
overview_stats <- data.frame(statistic = c("N Reads",
                                          "N UMIs",
                                          "N Genes",
                                          "Reads per UMI",
                                          "UMIs per Gene",
                                          "Reads per Gene"),
                             q_25 = c(quantile(meta$n_reads, 0.25),
                                      quantile(meta$n_umis,  0.25),
                                      quantile(meta$n_genes, 0.25),
                                      quantile(meta$n_reads/meta$n_umis, 0.25, na.rm = TRUE),
                                      quantile(meta$n_umis/meta$n_genes, 0.25, na.rm = TRUE),
                                      quantile(meta$n_reads/meta$n_genes, 0.25, na.rm = TRUE)),
                             median = c(quantile(meta$n_reads, 0.5),
                                        quantile(meta$n_umis,  0.50),
                                        quantile(meta$n_genes, 0.50),
                                        quantile(meta$n_reads/meta$n_umis, 0.50, na.rm = TRUE),
                                        quantile(meta$n_umis/meta$n_genes, 0.50, na.rm = TRUE),
                                        quantile(meta$n_reads/meta$n_genes, 0.50, na.rm = TRUE)),
                             q_75 = c(quantile(meta$n_reads, 0.75),
                                      quantile(meta$n_umis,  0.75),
                                      quantile(meta$n_genes, 0.75),
                                      quantile(meta$n_reads/meta$n_umis, 0.75, na.rm = TRUE),
                                      quantile(meta$n_umis/meta$n_genes, 0.75, na.rm = TRUE),
                                      quantile(meta$n_reads/meta$n_genes, 0.75, na.rm = TRUE)))

overview_stats$q_25 <- round(overview_stats$q_25, 2)
overview_stats$median <- round(overview_stats$median, 2)
overview_stats$q_75 <- round(overview_stats$q_75, 2)

knitr::kable(overview_stats, caption = "Well Stats Overview")
```

Plot read/umi/gene histograms
```{r QC histograms}
reads_plot <- qc_hist_plot(meta,
                           column = "n_reads",
                           name_x = "N Reads per Cell",
                           fill = "dodgerblue",
                           target = 2e4) +
  ggtitle("Alignment/Mapping Distributions")

umis_plot <- qc_hist_plot(meta,
                           column = "n_umis",
                           name_x = "N UMIs per Cell",
                           fill = "purple",
                           target = 8e3)

genes_plot <- qc_hist_plot(meta,
                           column = "n_genes",
                           name_x = "N Genes per Cell",
                           fill = "orangered",
                           target = 2e3)

histogram_list <- list(reads_plot,
                       umis_plot,
                       genes_plot)

plot_grid(plotlist = histogram_list,
          ncol = 1,
          nrow = 3)
```

Plot read/umi/gene scatters
```{r Read UMI Gene Scatter Plots}
qc_scatter_plot(meta,
                column_x = "n_reads",
                name_x = "N Reads per Cell",
                column_y = "n_umis",
                name_y = "N UMIs per Cell",
                color = "darkblue") +
  ggtitle("UMIs by Reads")

qc_scatter_plot(meta,
                column_x = "n_umis",
                name_x = "N UMIs per Cell",
                column_y = "n_genes",
                name_y = "N Genes per Cell",
                color = "red") +
  ggtitle("Genes by UMIs")

qc_scatter_plot(meta,
                column_x = "n_reads",
                name_x = "N Reads per Cell",
                column_y = "n_genes",
                name_y = "N Genes per Cell",
                color = "darkgreen") +
  ggtitle("Genes by Reads")
```

Metadata prep
```{r}
meta <- as.data.table(meta)
if("pbmc_sample_id" %in% names(meta)) {
  meta$plot_barcode <- paste0(meta$hto_barcode,"\n",meta$pbmc_sample_id)
} else {
  meta$pbmc_sample_id <- "No PBMC ID"
  meta$plot_barcode <- meta$hto_barcode
}
```


Show read/umi/gene stats per HTO category
```{r HTO Category Read UMI Gene Stats}
hto_order <- c("no_hash","singlet","doublet","multiplet")

hto_count_stats <- meta[, .(n_cells = nrow(.SD),
                            frac_cells = round(nrow(.SD)/nrow(meta), 4),
                            total_aligned_reads = sum(.SD$n_reads),
                            frac_all_aligned_reads = round(sum(.SD$n_reads)/sum(meta$n_reads), 4)),
                        by = hto_category]
hto_count_stats <- hto_count_stats[match(hto_order,hto_count_stats$hto_category),]
knitr::kable(hto_count_stats, caption = "HTO Category Count Stats")

hto_read_stats <- meta[, .(q_25 = quantile(n_reads, 0.25),
                      median = quantile(n_reads, 0.50),
                      q_75 = quantile(n_reads, 0.75)),
                  by = hto_category]
hto_read_stats <- hto_read_stats[match(hto_order,hto_read_stats$hto_category),]
knitr::kable(hto_read_stats, caption = "HTO Category Read Stats")

hto_umi_stats <- meta[, .(q_25 = quantile(n_umis, 0.25),
                      median = quantile(n_umis, 0.50),
                      q_75 = quantile(n_umis, 0.75)),
                  by = hto_category]
hto_umi_stats <- hto_umi_stats[match(hto_order,hto_umi_stats$hto_category),]
knitr::kable(hto_umi_stats, caption = "HTO Category UMI Stats")


hto_gene_stats <- meta[, .(q_25 = quantile(n_genes, 0.25),
                      median = quantile(n_genes, 0.50),
                      q_75 = quantile(n_genes, 0.75)),
                  by = hto_category]
hto_gene_stats <- hto_gene_stats[match(hto_order,hto_gene_stats$hto_category),]
knitr::kable(hto_gene_stats, caption = "HTO Category Gene Stats")
```

Cells per category barplot:
```{r Category Count Barplot}
hto_count_stats$fill <- c("black",
                          "darkblue",
                          "dodgerblue",
                          "mediumorchid4")

ggplot() +
  geom_bar(data = hto_count_stats,
           aes(x = as.factor(1:nrow(hto_count_stats)),
               y = n_cells,
               fill = fill),
           stat = "identity") +
  geom_text(data = hto_count_stats,
            aes(x = as.factor(1:nrow(hto_count_stats)),
                y = n_cells + 0.02 * max(n_cells),
                label = n_cells),
            hjust = 0.5,
            vjust = 0) +
  scale_fill_identity() +
  scale_x_discrete("",
                   breaks = 1:nrow(hto_count_stats),
                   labels = hto_count_stats$hto_category) +
  scale_y_continuous("N Cells", limits = c(0, 4e4)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 0,
                                   vjust = 0.3)) +
  ggtitle("Cells per HTO Category")
```

Category fraction barplot:
```{r Category Fraction Barplot}
plot_df <- hto_count_stats
plot_df$cells_max <- cumsum(plot_df$frac_cells)
plot_df$cells_min <- shift(plot_df$cells_max, type = "lag", fill = 0)
plot_df$reads_max <- cumsum(plot_df$frac_all_aligned_reads)
plot_df$reads_min <- shift(plot_df$reads_max, type = "lag", fill = 0)

ggplot() +
  geom_rect(data = plot_df,
            aes(xmin = reads_min,
                xmax = reads_max,
                ymin = 0.6,
                ymax = 1.4,
                fill = fill)) +
  geom_rect(data = plot_df,
            aes(xmin = cells_min,
                xmax = cells_max,
                ymin = 1.6,
                ymax = 2.4,
                fill = fill)) +
  scale_fill_identity() +
  scale_x_continuous("") +
  scale_y_continuous("",
                     breaks = c(1, 2),
                     labels = c("Fraction of Reads",
                                "Fraction of Cell Barcodes")) +
  theme_bw() +
  ggtitle("Fraction of Cells and Reads per HTO Category")
```


Show read/umi/gene stats per singlet hash
```{r HTO Barcode Read UMI Gene Stats}
singlet_meta <- as.data.table(meta[meta$hto_category == "singlet",])

bc_count_stats <- singlet_meta[, .(n_singlets = nrow(.SD),
                                   frac_singlets = round(nrow(.SD)/nrow(singlet_meta), 4)),
                               by = list(pbmc_sample_id,hto_barcode)]
knitr::kable(bc_count_stats, caption = "HTO Barcode Count Stats")

bc_read_stats <- singlet_meta[, .(q_25 = quantile(n_reads, 0.25),
                                  median = quantile(n_reads, 0.50),
                                  q_75 = quantile(n_reads, 0.75)),
                              by = list(pbmc_sample_id,hto_barcode)]

knitr::kable(bc_read_stats, caption = "HTO Barcode Singlet Read Stats")

bc_umi_stats <- singlet_meta[, .(q_25 = quantile(n_umis, 0.25),
                                 median = quantile(n_umis, 0.50),
                                 q_75 = quantile(n_umis, 0.75)),
                             by = list(pbmc_sample_id,hto_barcode)]
knitr::kable(bc_umi_stats, caption = "HTO Barcode Singlet UMI Stats")


bc_gene_stats <- singlet_meta[, .(q_25 = quantile(n_genes, 0.25),
                                  median = quantile(n_genes, 0.50),
                                  q_75 = quantile(n_genes, 0.75)),
                              by = list(pbmc_sample_id,hto_barcode)]
knitr::kable(bc_gene_stats, caption = "HTO Barcode Singlet Gene Stats")
```

Cells per barcode barplot:
```{r BC Count Barplot}
bc_to_plot_bc <- data.frame(hto_barcode = unique(singlet_meta$hto_barcode),
                            plot_barcode = unique(singlet_meta$plot_barcode))

bc_count_stats <- bc_count_stats[bc_to_plot_bc,
                                 on = "hto_barcode"]

ggplot() +
  geom_bar(data = bc_count_stats,
           aes(x = as.factor(plot_barcode),
               y = n_singlets),
           stat = "identity") +
  geom_hline(aes(yintercept = median(bc_count_stats$n_singlets)),
             linetype = "dashed") +
  scale_x_discrete("") +
  scale_y_continuous("N Cells", 
                     limits = c(0, 4000)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90,
                                   hjust = 1,
                                   vjust = 0.3)) +
  ggtitle("Cells per HTO Barcode")
```


Read distributions by HTO category and HTO barcode
```{r}
category_reads_violins <- qc_violin_plot(meta,
                                         category_x = "hto_category",
                                         name_x = "HTO Category",
                                         column_y = "n_reads",
                                         name_y = "N Reads per Cell",
                                         fill = "dodgerblue")

barcode_reads_violins <- qc_violin_plot(meta[meta$hto_category == "singlet",],
                                        category_x = "plot_barcode",
                                        name_x = "HTO Barcode (singlets)",
                                        column_y = "n_reads",
                                        name_y = "N Reads per Cell",
                                        fill = "dodgerblue") +
  ggtitle("Reads per Cell")

reads_violin_list <- list(category_reads_violins, 
                          barcode_reads_violins)

plot_grid(plotlist = reads_violin_list,
          ncol = 2, rel_widths = c(1, 3),
          nrow = 1, align = "h")
```

UMI distributions by HTO category and HTO barcode
```{r}
category_umis_violins <- qc_violin_plot(meta,
                                         category_x = "hto_category",
                                         name_x = "HTO Category",
                                         column_y = "n_umis",
                                         name_y = "N UMIs per Cell",
                                         fill = "purple")

barcode_umis_violins <- qc_violin_plot(meta[meta$hto_category == "singlet",],
                                        category_x = "plot_barcode",
                                        name_x = "HTO Barcode (singlets)",
                                        column_y = "n_umis",
                                        name_y = "N UMIs per Cell",
                                        fill = "purple") +
  ggtitle("UMIs per Cell")

umis_violin_list <- list(category_umis_violins, 
                          barcode_umis_violins)

plot_grid(plotlist = umis_violin_list,
          ncol = 2, rel_widths = c(1, 3),
          nrow = 1, align = "h")
```

Gene distributions by HTO category and HTO barcode
```{r}
category_genes_violins <- qc_violin_plot(meta,
                                         category_x = "hto_category",
                                         name_x = "HTO Category",
                                         column_y = "n_genes",
                                         name_y = "N Genes per Cell",
                                         fill = "orangered")

barcode_genes_violins <- qc_violin_plot(meta[meta$hto_category == "singlet",],
                                        category_x = "plot_barcode",
                                        name_x = "HTO Barcode (singlets)",
                                        column_y = "n_genes",
                                        name_y = "N Genes per Cell",
                                        fill = "orangered") +
  ggtitle("Genes per Cell")

genes_violin_list <- list(category_genes_violins, 
                          barcode_genes_violins)

plot_grid(plotlist = genes_violin_list,
          ncol = 2, rel_widths = c(1, 3),
          nrow = 1, align = "h")
```

Split list based on hashes
```{r Split H5 Lists}
stm("Splitting HDF5 based on Hash Category Table")
h5_list <- split_h5_list_by_hash(h5_list)
```

Write HDF5 files
```{r Write split files}
for(i in seq_along(h5_list)) {
  out_file <- file.path(out_dir, paste0(in_well, "_", names(h5_list)[i],".h5"))
  stm(paste0("Writing HDF5 to ", out_file))
  write_h5_list(h5_list[[i]],
                h5_file = out_file)
  h5closeAll()
}
```

```{r Session Info}
sessionInfo()
```

Total time elapsed
```{r Show Time}
end_time <- Sys.time()
diff_time <- end_time - start_time
time_message <- paste0("Elapsed Time: ", 
                       round(diff_time, 3),
                       " ", units(diff_time))
print(time_message)
stm(time_message)
stm("HTO count processing complete.")
```
